Analysis
========

Project startup
---------------

Setting knitr options.

``` {r knitr_options}
require(knitr)
opts_knit$set(root.dir="..")
opts_chunk$set(fig.path="../graphs/analysis-")
```

Load required libraries.
```{r load_libraries, warning=FALSE}
require(Hmisc)
require(RColorBrewer)
```
Load project with ProjectTemplate.

```{r load_project, warning=FALSE}
require(ProjectTemplate)
reload.project()
```

Load munge R script.
```{r external_code}
read_chunk("munge/01-A.R")
```

Data Munging
------------

Based on initial data inspection and exploration data were tidied up and some
columns were added to help the analysis.

Here the most significant munging.

### Compute FICO.Range median values

FICO score is the principal predictor of loan rate.

To better analyse this relation, a column was added with FICO.Range median value.
Values of this column were obtained transforming from character to numeric the
lower and upper limits of each range.
```{r, eval=FALSE}
<<fico_lower_upper_median>>
```
```{r, echo=FALSE}
names(table(loansData$FICO.Median))
```

### Cut Amount.Funded.By.Investors in ranges

Exploration revelead the amount funded is a confounder for the relation between
interest rate and FICO score.

To better analyse the influence of this confounder, a column was added with ranges
of amount funded with the cut2 function (3 percentiles).
```{r, eval=FALSE}
<<amount_funded_ranges>>
```

Previously amount funded values were tidied, eliminating 0 and -0.10 values
(considered not significant for the analysis) and substituting them with NAs.
```{r, eval=FALSE}
<<amount_funded_zero2na>>
```

### Preparing data frame for the analysis

The data frame used for exploration consisted in 2500 observation on 18 variables.

It is subsetted to:
*  mantain only the variables used for the analysis (6)
*  filter the observations to exclude those with amount funded equal to NA (2494 remaining)

```{r data_subset}
loansDataAn <- loansData[,c(3,2,18,4,10,17)]
loansDataAn <- loansDataAn[!is.na(loansDataAn$Amount.Funded.By.Investors),]
```

Finally Amount Funded range, FICO range, and Loand Lenghts are converted to factor.
```{r range2factor}
loansDataAn$Amount.Funded.Range <- as.factor(loansDataAn$Amount.Funded.Range)
loansDataAn$FICO.Range <- as.factor(loansDataAn$FICO.Range)
loansDataAn$Loan.Length.Months <- as.factor(loansDataAn$Loan.Length.Months)
```

Data Summaries
--------------

```{r data_summaries}
dim(loansDataAn)
head(loansDataAn)
summary(loansDataAn)
```

Analyses
--------

### Interest rate and its predictors

#### FICO score

There is a clear negative correlation between loan rate and FICO score.

```{r rate_fico_boxplot, fig.height=5, fig.width=10}
boxplot(loansDataAn$Interest.Rate ~ loansDataAn$FICO.Range, xaxt = "n")
loansByFICORange <- table(loansDataAn$FICO.Range)
axis(1, las=2, at = 1:nrow(loansByFICORange), labels = names(loansByFICORange))
```
```{r rate_fico_summary}
tapply(loansDataAn$Interest.Rate, loansDataAn$FICO.Range, summary)
```

#### Loan length

There is a slightly positive correlation between loan rate and loan length:
longer loans (60 months vs. 36 months) are risky and so interest rates are higher.
```{r rate_length_boxplot, fig.height=5, fig.width=5}
boxplot(loansDataAn$Interest.Rate ~ loansDataAn$Loan.Length.Months)
```
```{r rate_length_summary}
tapply(loansDataAn$Interest.Rate, loansDataAn$Loan.Length.Months, summary)
```

#### Amount funded

There is even a slightly positive correlation between loan rate and amount funded.
The range from $200 to $7,500 and from $7,500 to $14,075 have a similar median rate.
But the range from $14,075 to $35,000 has a median rate more than 2% higher.

```{r rate_funded_boxplot, fig.height=5, fig.width=7.5}
boxplot(loansDataAn$Interest.Rate ~ loansDataAn$Amount.Funded.Range)
```
```{r rate_funded_summary}
tapply(loansDataAn$Interest.Rate, loansDataAn$Amount.Funded.Range, summary)
```

### Models

#### Interest rate and FICO score (Model 1)

```{r model1_lm_summary}
lm1 <- lm(loansDataAn$Interest.Rate ~ loansDataAn$FICO.Median)
summary(lm1)
```
```{r model1_confint_summary}
confint(lm1, level=0.95)
```

```{r model1_plot, fig.height=7, fig.width=12}
rgbcols = c("#FF000033","#00FF0033","#0000FF33")
par(mfrow = c(1, 2), mar=c(5,5,1,1), oma=c(0,0,4,1))
# first panel
palette(rgbcols)
plot(loansDataAn$FICO.Median
     ,loansDataAn$Interest.Rate
     ,pch = 19
     ,col = loansDataAn$Loan.Length.Months
     ,ylim=c(0,max(loansDataAn$Interest.Rate))
     ,xlim=c(min(loansDataAn$FICO.Median),max(loansDataAn$FICO.Median))
     ,frame.plot=FALSE
     ,xlab="FICO score with Loan length", ylab="Interest Rate (%)")
lines(loansDataAn$FICO.Median, lm1$fitted,col="#70707099",lwd=3)
labels <- c("36 months","60 months")
position <- 'topright'
inset <- c(0.05, 0.05)
legend(position, labels, fill=rgbcols, inset=inset, bty = "n")
mtext(text="(a)",side=3,line=0)
# second panel
par(mar=c(5,1,1,1))
plot(loansDataAn$FICO.Median
     ,loansDataAn$Interest.Rate
     ,pch = 19
     ,col = loansDataAn$Amount.Funded.Range
     ,ylim=c(0,max(loansDataAn$Interest.Rate))
     ,xlim=c(min(loansDataAn$FICO.Median),max(loansDataAn$FICO.Median))
     ,frame.plot=FALSE
     ,xlab="FICO score with Amount funded", ylab=""
     ,yaxt = "n")
lines(loansDataAn$FICO.Median, lm1$fitted,col="#70707099",lwd=3)
labels <- c("[     $200 -   $7,500)",
            "[  $7,500 - $14,075)",
            "[$14,075 - $35,000]")
position <- 'topright'
inset <- c(0.05, 0.05)
legend(position, labels, fill=rgbcols, inset=inset, bty = "n")
mtext(text="(b)",side=3,line=0)
mtext("Figure 1. Lending Club loans. Relationship between Interest rate and FICO score.", side=3, line=1, outer=TRUE, font=2)
```
```{r model1_plot2, fig.height=8, fig.width=12}
par(mfrow = c(1, 3))
#palette(rev(colorRampPalette(brewer.pal(9,"Blues"))(40)))
palette(rev(colorRampPalette(rgbcols)(40)))
plot(loansDataAn$Interest.Rate, lm1$residuals, pch = 19, col = loansDataAn$FICO.Range)
palette(c(brewer.pal(5,"PRGn")[1],brewer.pal(5,"PRGn")[5]))
plot(loansDataAn$Interest.Rate, lm1$residuals, pch = 19, col = loansDataAn$Loan.Length.Months)
palette(c(brewer.pal(9,"Reds")[3],brewer.pal(9,"Reds")[6],brewer.pal(9,"Reds")[9]))
plot(loansDataAn$Interest.Rate, lm1$residuals, pch = 19, col = loansDataAn$Amount.Funded.Range)
```

#### Interest Rate, FICO score and Loan length (Model 2)

```{r model2_lm_summary}
lm2 <- lm(loansDataAn$Interest.Rate ~ loansDataAn$FICO.Median + loansDataAn$Loan.Length.Months)
summary(lm2)
```
```{r model2_confint_summary}
confint(lm2, level=0.95)
```
```{r model2_plot, fig.height=8, fig.width=12}
par(mfrow = c(1, 3))
#palette(rev(colorRampPalette(brewer.pal(9,"Blues"))(40)))
palette(rev(colorRampPalette(rgbcols)(40)))
plot(loansDataAn$Interest.Rate, lm2$residuals, pch = 19, col = loansDataAn$FICO.Range)
palette(c(brewer.pal(5,"PRGn")[1],brewer.pal(5,"PRGn")[5]))
plot(loansDataAn$Interest.Rate, lm2$residuals, pch = 19, col = loansDataAn$Loan.Length.Months)
palette(c(brewer.pal(9,"Reds")[3],brewer.pal(9,"Reds")[6],brewer.pal(9,"Reds")[9]))
plot(loansDataAn$Interest.Rate, lm2$residuals, pch = 19, col = loansDataAn$Amount.Funded.Range)
```

#### Interest Rate, FICO score, Loan length and Amount funded (Model 3)

```{r model3_lm_summary}
lm3 <- lm(loansDataAn$Interest.Rate ~ loansDataAn$FICO.Median + loansDataAn$Loan.Length.Months + loansDataAn$Amount.Funded.Range)
summary(lm3)
```
```{r model3_confint_summary}
confint(lm3, level=0.95)
```
```{r model3_plot, fig.height=8, fig.width=12}
par(mfrow = c(1, 3))
#palette(rev(colorRampPalette(brewer.pal(9,"Blues"))(40)))
palette(rev(colorRampPalette(rgbcols)(40)))
plot(loansDataAn$Interest.Rate, lm3$residuals, pch = 19, col = loansDataAn$FICO.Range)
palette(c(brewer.pal(5,"PRGn")[1],brewer.pal(5,"PRGn")[5]))
plot(loansDataAn$Interest.Rate, lm3$residuals, pch = 19, col = loansDataAn$Loan.Length.Months)
palette(c(brewer.pal(9,"Reds")[3],brewer.pal(9,"Reds")[6],brewer.pal(9,"Reds")[9]))
plot(loansDataAn$Interest.Rate, lm3$residuals, pch = 19, col = loansDataAn$Amount.Funded.Range)
```

Figure making
-------------

```{r figure}
## Save figure to pdf file
pdf(file="graphs/finalFigure.pdf", height=10, width=20, family="Helvetica", pointsize=18)
## Plot
rgbcols = c("#FF000033","#00FF0033","#0000FF33")
par(mfrow = c(1, 2), mar=c(5,6,1,.5), oma=c(0,0,4,1))
# first panel
palette(rgbcols)
plot(loansDataAn$FICO.Median
     ,loansDataAn$Interest.Rate
     ,pch = 19
     ,col = loansDataAn$Loan.Length.Months
     ,ylim=c(0,max(loansDataAn$Interest.Rate))
     ,xlim=c(min(loansDataAn$FICO.Median),max(loansDataAn$FICO.Median))
     ,frame.plot=FALSE
     ,xlab="FICO score with Loan length", ylab="Interest Rate (%)")
lines(loansDataAn$FICO.Median, lm1$fitted,col="#70707099",lwd=3)
labels <- c("36 months","60 months")
position <- 'topright'
inset <- c(0.05, 0.05)
legend(position, labels, fill=rgbcols, inset=inset, bty = "n")
mtext(text="(a)",side=3,line=0)
# second panel
par(mar=c(5,.5,1,6))
plot(loansDataAn$FICO.Median
     ,loansDataAn$Interest.Rate
     ,pch = 19
     ,col = loansDataAn$Amount.Funded.Range
     ,ylim=c(0,max(loansDataAn$Interest.Rate))
     ,xlim=c(min(loansDataAn$FICO.Median),max(loansDataAn$FICO.Median))
     ,frame.plot=FALSE
     ,xlab="FICO score with Amount funded",
     ,yaxt = "n")
axis(4)
mtext("Interest Rate (%)",side=4,line=3)
lines(loansDataAn$FICO.Median, lm1$fitted,col="#70707099",lwd=3)
labels <- c("[     $200 -   $7,500)",
            "[  $7,500 - $14,075)",
            "[$14,075 - $35,000]")
position <- 'topright'
inset <- c(0.05, 0.05)
legend(position, labels, fill=rgbcols, inset=inset, bty = "n")
mtext(text="(b)",side=3,line=0)
mtext("Figure 1. Lending Club loans. Relationship between Interest rate and FICO score.", side=3, line=1, outer=TRUE, font=2)
dev.off()
```
