Munge
=====

Tidying data before exploration and analysis.

Startup
-------

Setting knitr options.
``` {r knitr_options}
require(knitr)
opts_knit$set(root.dir="..")
opts_chunk$set(fig.path="../graphs/munge-")
```

Load required libraries.
```{r load_libraries, warning=FALSE}
require(Hmisc)
```
Load project with ProjectTemplate.
```{r load_project, warning=FALSE}
require(ProjectTemplate)
reload.project()
```

Load munge R script.
```{r external_code}
read_chunk("munge/01-A.R")
```

Interest.Rate and Debt.To.Income.Ratio, from character to numeric
-----------------------------------------------------------------
`Interest.Rate` and `Debt.To.Income.Ratio` are imported as character, but they are percentages.

Both are formatted with a percent sign:
```{r, echo=FALSE}
loansDataRaw[1:5, c(3,6)]
```

The percent signs are substituted away and the values are converted to numeric.
```{r, eval=FALSE}
<<percentage_to_numeric>>
```

Loan.Length, from character to integer
--------------------------------------
`Loan.Length` is imported as character, but it represents a time span in months.

It is formatted with the label "months":
```{r, echo=FALSE}
names(table(loansDataRaw$Loan.Length))
```

The "months" labels are substituted away, the values are converted to integer,
and the name changed to `Loan.Length.Months`.
```{r, eval=FALSE}
<<months_to_integer>>
```

Loan.Purpose, formatted with pascal case and dots
-------------------------------------------------
`Loan.Purpose` values are in lower case separated by underscores.
```{r, echo=FALSE}
names(table(loansDataRaw$Loan.Purpose))
```

Values are converted to pascal case (first letter of every word in upper case,
other letters in lower case), and undescores substituted wiht dots.
```{r, eval=FALSE}
<<purpose_to_pc>>
```

Home.Ownership, formatted with pascal case
------------------------------------------
`Home.Ownership` values are in upper case.
```{r, echo=FALSE}
names(table(loansDataRaw$Home.Ownership))
```

Values are converted to pascal case (first letter of every word in upper case,
other letters in lower case).
```{r, eval=FALSE}
<<ownership_to_pc>>
```

Employment.Length, from character to integer
---------------------------------------
`Employment.Length` is imported as character, but it represents a time span in years.
```{r, echo=FALSE}
names(table(loansDataRaw$Employment.Length))
```

The conversion steps are:

1.  substitute values representing less than one year (e.g. "< 1 year") with "0"
2.  remove unnecessary characters from the other values, leaving only the number (e.g. from "2 years" to "2");
    the value "10+ years", representing 10 or more years is substituted with 10
3.  convert all values to integer; missing values "n/a" are coerced to NA

```{r, eval=FALSE}
<<emplength_to_integer>>
```

Adding columns with FICO.Range lower, upper and median values
-------------------------------------------------------------
`FICO.Range` is imported as character, but it represents ranges of credit score.
```{r, echo=FALSE}
names(table(loansDataRaw$FICO.Range))
```

Conversion to numerical values could be useful for exploration and analysis.
Three columns are added, with the lower bound, the upper bound and the median
of the range.
```{r, eval=FALSE}
<<fico_lower_upper_median>>
```

Adding columns with Amount.Funded.By.Investors ranges
-----------------------------------------------------
`Amount.Funded.By.Investors` is imported as numeric.

Some dirty values are present: 0 and -0.10.
These are converted to NA.
```{r, eval=FALSE}
<<amount_funded_zero2na>>
```

Cutting into ranges could be useful for exploration and analysis.
One column is added, with three quantile ranges.
```{r, eval=FALSE}
<<amount_funded_ranges>>
```
